Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-17 paper2.qmd
section:appendix; process:write

@@

<!-- 1. What does empirical research do first? -->
Empirical research uses evidence from observation and experimentation to address real-world challenges. In this context, researchers typically formulate their research questions as *estimands* or *targets of inference*, i.e., the specific quantities they seek to determine [@Everitt_et_al_2010]. For instance, researchers might be interested in answering the following question: “To what extent do different teaching methods $(T)$ influence students' ability to produce high-quality written texts $(Y)$?" To investigate this, researchers could randomly assign students to two groups, each exposed to a different teaching method $(T_{i} = \{1,2\})$. Then, they would perform pairwise comparisons, generating a dichotomous outcome $(Y_{i} = \{0,1\})$ showing which student exhibits more of the ability. In this scenario, the research question can be rephrased as the estimand, “*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?” and this estimand can be mathematically represented by the random quantity $E[Y_{i} \: | \: T_{i}=1] - E[Y_{i} \: | \: T_{i}=2]$, where $E[\cdot]$ denotes the expected value.

@@

<!-- 3. But, what do we need to produce a good estimate? -->
While numerous methods can approximate an estimand, researchers prioritize estimators with desirable identification properties that ensure the accuracy of estimates. For instance, the Z-test is an estimator known for its effectiveness when comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006]. If this is the case, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$. The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i} \: | \: T_{i}=1]$ and $E[Y_{i} \: | \: T_{i}=2]$, respectively. The noise, represented by $\hat{s}_{p}$, is defined as the unpooled sample variability observed between the two groups.

@@

Randomized experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014]. This recognition stems from their ability to enable researchers to interpret associational estimates as causal. They achieve this by ensuring data, and by extension an estimator, satisfies several key properties, such as common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020]. The most critical property, however, is the elimination of confounding. *Confounding* occurs when an external variable $X$ simultaneously influences the outcome $Y$ and the variable of interest $T$, resulting in spurious associations [@Everitt_et_al_2010]. Randomization addresses this issue by decoupling the association between the intervention allocation $T$ from any other variable $X$ [@Morgan_et_al_2014; @Neal_2020].

@@

<!-- 2. What are counterfactuals? -->
*Counterfactuals* are hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a given cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. The structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] provides a formal framework for defining counterfactuals. For instance, in the scenario described in @sec-appendix-A, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes: $\tau_{i} = Y_{i} \: | \: do(T_{i}=1) - Y_{i} \: | \: do(T_{i}=2)$. Here, $do(T_{i}=t)$ represents the intervention operator, $Y_{i} \: | \: do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i} \: | \: do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$. Note that if a student is assigned to intervention $T_{i}=1$, the potential outcome under $T_{i}=2$ becomes a counterfactual, as it is no longer observed nor observable. To address the challenge of unobserved counterfactuals, the approach extends the ICE to the *average causal effect* (ACE): $\tau = E[\tau_{i}] = E[Y_{i} \: | \: do(T_{i}=1)]- E[Y_{i} \: | \: do(T_{i}=2)]$, representing the average difference between observed potential outcomes and their counterfactual counterparts.

@@

<!-- 3. But how does it work? -->
Even when data originates from an observational study, researchers can identify the ACE from associational estimates using the structural approach. They achieve this by performing statistical conditioning on a *sufficient adjustment set* of variables $X$ [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. This *sufficient* set (potentially empty) must block all non-causal paths between $T$ to $Y$ without opening new ones, ensuring the ACE estimator satisfies several key properties, including confounding elimination. If such a set exists, then $T$ and $Y$ are *d-separated* by $X$ [@Pearl_2009], meaning researchers can estimate the ACE from associational random quantities [@Morgan_et_al_2014]. Naturally, the validity of claims about the effects of $T$ on $Y$ now hinges on the assumption that $X$ serves as a sufficient adjustment set. However, as @Kohler_et_al_2019 [pp. 150] observed, drawing conclusions about the real world from observed data inevitably requires assumptions. This holds true for both observational and experimental data.

@@

For instance, if researchers are unable to conduct the randomized experiments described in @sec-appendix-A and must rely on observational data, they can still estimate the ACE, provided that a variable $X$, such as the socio-economic status of the school, blocks all non-causal paths between the teaching method $T$ and the outcome $Y$ without opening any new ones. Under these conditions, the ACE can be estimated from the *conditional average causal effect* (CACE), defined by $E[Y_{i} \: | \: T_{i}=t, X]$. Using the CACE, the ACE can be expressed as $\tau =$ $E[Y_{i} \: | \: do(T_{i}=1)] - E[Y_{i} \: | \: do(T_{i}=2)] =$ $E_{X}\left[ E[Y_{i} \: | \: T_{i}=1, X] - E[Y_{i} \: | \: T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014]. Notably, the approach extends the ACE for a continuous variable $T$ as $\tau =$ $E[Y_{i} \: | \: do(T_{i}=t)] =$ $d E_{X}\left[ E[Y_{i} \: | \: T_{i}=t, X]\right]/ dt$, ensuring broad applicability across different causal scenarios [@Neal_2020, pp. 45].

@@

<!-- commands for d-separation -->
\newcommand{\dsep}{\perp\!\!\!\perp}
<!-- \newcommand{\ndsep}{\not\!\perp\!\!\!\perp} -->

<!-- 1. The benefits of SCMs and DAGs -->
The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]. Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that help researchers determine which statistical models can identify an estimand, assuming the depicted causal structure is correct, thus enabling valid causal inference [@McElreath_2020]. They achieve this by representing various causal structures, regardless of complexity, using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Furthermore, SCMs and DAGs depict causal relationships in a non-parametric and fully interactive way. This flexibility enables feasible ACE identification strategies without requiring detailed specifications of functional relationships relating the variables, their parameters, or their data types [@Pearl_et_al_2016]. @fig-IEflow shows the role of theoretical models in the inference process.

@@

<!-- 4. Theoretical implications of the basic building block of the SCMS and DAGs -->
A careful examination of the figures highlights the theoretical assumptions underlying these building blocks. SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X_{1}$ and $X_{3}$ are independent or not causally related. SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2] illustrate two connected nodes, representing a scenario where a *parent* node $X_{1}$ exerts a causal influence on a *child* node $X_{3}$. In this setup, $X_{3}$ is considered a *descendant* of $X_{1}$. Additionally, $X_{1}$ and $X_{3}$ are described as *adjacent* because an edge directly connects them. SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] depict a *chain*, where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$. In this configuration, $X_{1}$ is a parent node of $X_{2}$, which is a parent node of $X_{3}$. This structure creates a *directed path* between $X_{1}$ and $X_{3}$. Consequently, $X_{1}$ is an *ancestor* of $X_{3}$, and $X_{2}$ fully *mediates* the relationship between the two. SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4] illustrate a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$. Here, $X_{2}$ is a parent node that *confounds* the relationship between $X_{1}$ and $X_{3}$. Finally, SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] show a *collider*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$. In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$ (an *inmorality*). Notably, all building blocks assume the errors are independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X1} \dsep e_{X2}$, $e_{X1} \dsep e_{X3}$, and $e_{X2} \dsep e_{X3}$.

@@

## The probabilistic and causal implications of SCMs and DAGs {#sec-appendix-D .appendix appendix-style=plain}

<!-- 1. How the SCMS and DAGs encode probabilistic information? -->
Beyond their graphical capabilities, SCMs and DAGs can encode the probabilistic information embedded within a causal structure. They achieve this encoding by relying on three fundamental assumptions: the local Markov, the minimality, the causal edges and the modularity assumptions. The *local Markov assumption* encodes probabilistic independencies between variables by declaring that nodes in a graph are independent of all its non-descendants, given its parents [@Neal_2020, pp. 20]. Meanwhile, the *minimality assumption* encodes probabilistic dependencies among variables by stating that every pair of adjacent nodes exhibits a dependency [@Neal_2020, pp. 21]. Finally, the *causal edges assumption* encodes causal relationships between variables by declaring that each parent node acts as a direct cause of its children [@Neal_2020, pp. 22]. @fig-ACflow illustrates how these assumptions influence the statistical and causal interpretations of graphs.

@@

<!-- 2. What are the general implications of this? -->
The encoding of probabilistic information in SCMs and DAGs facilitates the efficient representation of the joint distribution of variables [@Pearl_et_al_2016, pp. 29]. A notable implication of the assumptions underlying the probabilistic encoding is that any conceptual model described by an SCM and DAG can express the joint distribution of variables as a product of conditional probability distributions (CPDs) in the form $P(child \: | \: parents)$. This property is formally known as the *Bayesian Network factorization* (BNF) [@Pearl_et_al_2016, pp. 29; @Neal_2020, pp. 21]. In this factorization, depicted in @eq-net_factor, $pa(X_{i})$ denotes the set of variables that are the parents of $X_{i}$.

$$
P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \: | \: pa(X_{i}) )
$$ {#eq-net_factor}

<!-- 3. What is the net result of this? -->
This encoding enables researchers with conceptual (theoretical) knowledge in the form of an SCM and DAG to predict patterns of (in)dependencies in the data. As highlighted by @Pearl_et_al_2016 [pp. 35], these predictions depend solely on the structure of these conceptual models without requiring the quantitative details of the equations or the distributions of the errors. Moreover, once researchers observe empirical data, the patterns of (in)dependencies in this data can provide significant insights into the validity of the proposed conceptual model. Moreover, it allows to express

@@

<!-- 4. can you illustrate this concept? -->
The five fundamental building blocks described in @sec-appendix-C illustrate how to predict (in)dependencies in data. For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$). This conclusion follows from expressing their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$, which defines probabilistic independence [@Neal_2020, pp. 24]. Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$, as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \: | \: X_{1}) P(X_{1})$. Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in the figures of @sec-appendix-C and the CPDs.

<!-- ($X_{1} \ndsep X_{3}$) -->

Beyond the insights gained from two-node structures, researchers can uncover more nuanced patterns of (in)dependencies from chains, forks, and colliders. By leveraging these patterns, researchers can identify the ACE from associational estimates through statistical conditioning, thereby disentangling causation from mere association. For instance, researchers applying the BNF to the chain structure depicted in the SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] can represent the joint distribution for the observed variables as $P(X_{1},X_{2},X_{3}) = P(X_{1}) P(X_{2} \: | \: X_{1}) P(X_{3} \: | \: X_{2})$. From this expression, researchers can infer additional (in)dependencies assumed in any data generated by this structure. Specifically, the structure implies that $X_{1}$ and $X_{3}$ are unconditionally dependent, but conditionally independent when controlling for $X_{2}$ ($X_{1} \dsep X_{3} \: | \: X_{2}$) [@Pearl_et_al_2016, pp. 37; @Neal_2020, pp. 25]. In the fork structure shown in the SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4], researchers can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) = P(X_{1} \: | \: X_{2}) P(X_{2}) P(X_{3} \: | \: X_{2})$. This expression allows researchers to further infer that $X_{1}$ and $X_{3}$ are unconditionally dependent, but conditionally independent when controlling for $X_{2}$ ($X_{1} \dsep X_{3}  \: | \: X_{2}$) [@Pearl_et_al_2016, pp. 40]. Finally, researchers analyzing the collider structure illustrated in the SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) = P(X_{1}) P(X_{2} \: | \: X_{1}, X_{3}) P(X_{3})$. This representation allows researchers to infer that $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$), but conditionally dependent when controlling for $X_{2}$ [@Pearl_et_al_2016, pp. 41; @Neal_2020, pp. 26].

<!-- ($X_{1} \ndsep X_{3}$) -->
<!-- ($X_{1} \ndsep X_{3}$) -->
<!-- ($X_{1} \ndsep X_{3} \: | \: X_{2}$)  -->

<!-- 5. But, how do we identify causal effects?  -->
Ultimately, after encoding probabilistic information, researchers can use SCMs and DAGs to identify causal effects. This process relies on one additional assumption: the *modularity assumption*. This assumption asserts that intervening on a node only alters its causal mechanism, implying that interventions are local [@Neal_2020, pp. 34]. From this assumption, concepts such as manipulated DAGs and the Truncated factorization emerge. *Manipulated graphs* are DAGs with removed edges, signifying intervention. Similarly, the *Truncated factorization* represents these manipulated DAGs within SCMs (@eq-trunc_factor). In this expression, $S$ denotes the subset of manipulated variables $X_{p}$.

$$
\begin{aligned}
P(X_{1}, X_{2}, \dots, X_{P} \: | \: do(S=s)) & = \prod_{p \not\in S} P(X_{i} \: | \: pa(X_{i}) ) \\
P(X_{1}, X_{2}, \dots, X_{P} \: | \: do(S=s)) & = 1 \quad \quad (\text{otherwise})
\end{aligned}
$$ {#eq-trunc_factor}

<!-- 6. Continuing with the motivating example -->
Researchers can apply these new insights to revisit the scenario in @sec-appendix-B. From the SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1a], they can infer that the effect of $T$ on $Y$ is confounded by $X$. This is evident because the effect decomposes into a direct causal effect ($T \rightarrow Y$, black path) and a non-causal effect arising from the confounder $X$ ($T \leftarrow X \rightarrow Y$, gray path). Assuming the causal structure is correct, researchers can identify the CACE of $T$ on $Y$ by conditioning on $X$ and then marginalize this effect over $X$ to obtain the ACE. Conditioning on $X$ means restricting the focus to schools with a specific socio-economic status $X$ [@Neal_2020, pp. 32]. Notably, researchers often depict conditioning as a square around a variable. Furthermore, the conditioning process leads to the manipulated DAG [-@fig-mdag_example1b], equivalent to the conditioned DAG [-@fig-mdag_example1a].

>>>>

<!-- commands for d-separation -->
\newcommand{\dsep}{\:\bot\:}
\newcommand{\ndsep}{\:\not\bot\:}
\newcommand{\cond}{\:|\:}

@@

<!-- 1. What does empirical research do first? -->
Empirical research uses evidence from observation and experimentation to address real-world challenges. In this context, researchers typically formulate their research questions as *estimands* or *targets of inference*, i.e., the specific quantities they seek to determine [@Everitt_et_al_2010]. For instance, researchers might be interested in answering the following question: “To what extent do different teaching methods $(T)$ influence students' ability to produce high-quality written texts $(Y)$?" To investigate this, researchers could randomly assign students to two groups, each exposed to a different teaching method $(T_{i} = \{1,2\})$. Then, they would perform pairwise comparisons, generating a dichotomous outcome $(Y_{i} = \{0,1\})$ showing which student exhibits more of the ability. In this scenario, the research question can be rephrased as the estimand, “*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?” and this estimand can be mathematically represented by the random associational quantity in @eq-group_diff, where $E[\cdot]$ denotes the expected value.

$$
E[Y_{i} \cond T_{i}=1] - E[Y_{i} \cond T_{i}=2]
$$ {#eq-group_diff}

@@

<!-- 3. But, what do we need to produce a good estimate? -->
While numerous methods can approximate an estimand, researchers prioritize estimators with desirable identification properties that ensure the accuracy of estimates. For instance, the Z-test is an estimator known for its effectiveness when comparing groups' proportions, yielding accurate estimates when the underlying assumptions of the statistic are met [@Kanji_2006]. If this is the case, the Z-test is expressed as a signal-to-noise statistic $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$. The signal is defined as the difference between the groups' sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y_{i} \cond T_{i}=1]$ and $E[Y_{i} \cond T_{i}=2]$, respectively. The noise, represented by $\hat{s}_{p}$, is defined as the unpooled sample variability observed between the two groups.

@@

Randomized experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014]. This recognition stems from their ability to enable researchers interpret associational estimates as causal. They achieve this by ensuring data, and by extension an estimator, satisfies several key properties, such as common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020]. The most critical property, however, is the elimination of confounding. *Confounding* occurs when an external variable $X$ simultaneously influences the outcome $Y$ and the variable of interest $T$, resulting in spurious associations [@Everitt_et_al_2010]. Randomization addresses this issue by decoupling the association between the intervention allocation $T$ from any other variable $X$ [@Morgan_et_al_2014; @Neal_2020].

@@

<!-- 2. What are counterfactuals? -->
*Counterfactuals* are hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a given cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. The structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] provides a formal framework for defining counterfactuals. For instance, in the scenario described in @sec-appendix-A, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes, as in @eq-ICE.

$$
\tau_{i} = Y_{i} \cond do(T_{i}=1) - Y_{i} \cond do(T_{i}=2)
$$ {#eq-ICE}

where $do(T_{i}=t)$ represents the intervention operator, $Y_{i} \cond do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i} \cond do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$. Here, an *intervention* involves assigning a constant value to the treatment variable for each student's potential outcomes. In contrast, conditioning involves restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp. 32] (see @eq-group_diff).  
<!-- Conditioning on T=t just means that we are restricting our focus to the subset of the population to those who received treatment t. In contrast, an intervention would be to take the whole population and give everyone treatment t [@Neal_2020, pp. 32]. -->

Note that if a student is assigned to intervention $T_{i}=1$, the potential outcome under $T_{i}=2$ becomes a counterfactual, as it is no longer observed nor observable. To address this challenge, the structural approach extends the ICE to the *average causal effect* (ACE, @eq-ACE), representing the average difference between observed potential outcomes and their counterfactual counterparts.

$$
\tau = E[\tau_{i}] = E[Y_{i} \cond do(T_{i}=1)]- E[Y_{i} \cond do(T_{i}=2)]
$$ {#eq-ACE}

@@

<!-- 3. But how does it work? -->
Even though counterfactuals are unobserved and unobservable, researchers can still identify the ACE from associational estimates by leveraging the structural approach. The approach identifies the ACE by statistically conditioning data on a *sufficient adjustment set* of variables $X$ [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. This *sufficient* set (potentially empty) must block all non-causal paths between $T$ to $Y$ without opening new ones. If such a set exists, then $T$ and $Y$ are *d-separated* by $X$ ($T \dsep Y \cond X$) [@Pearl_2009], and $X$ satisfies the *backdoor criterion* [@Neal_2020, pp 37].

Conditioning on a sufficient adjustment set enables researchers to estimate the ACE, even when the data comes from an observational study. This process is feasible because such conditioning ensures that the ACE estimator satisfies several critical properties, including confounding elimination [@Morgan_et_al_2014]. Naturally, the validity of claims about the causal effects of $T$ on $Y$ now hinges on the assumption that $X$ serves as a sufficient adjustment set. However, as @Kohler_et_al_2019 [pp. 150] noted, drawing conclusions about the real world from observed data inevitably requires assumptions. This requirement holds true for both observational and experimental data.

@@

For instance, If researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as a variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion. Under these circumstances, researchers firt identify the *conditional average causal effect* (CACE, @eq-CACE)

$$
E[Y_{i} \cond T_{i}=t, X]
$$ {#eq-CACE}

From the CACE, they identify the ACE from associational quantities as in @eq-mACE. Here, $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].

$$
\begin{aligned}
  \tau & = E[Y_{i} \cond do(T_{i}=1)] - E[Y_{i} \cond do(T_{i}=2)] \\
  & = E_{X}\left[ E[Y_{i} \cond T_{i}=1, X] - E[Y_{i} \cond T_{i}=2, X] \right]
\end{aligned}
$$ {#eq-mACE}

Notably, the approach extends the ACE identification for a continuous variable $T$ as in @eq-mACE_cont, ensuring broad applicability across different causal scenarios [@Neal_2020, pp. 45]

$$
\begin{aligned}
  \tau &= E[Y_{i} \cond do(T_{i}=t)] \\
  & = d E_{X}\left[ E[Y_{i} \cond T_{i}=t, X]\right]/ dt
  \end{aligned}
$$ {#eq-mACE_cont}

@@

<!-- 1. The benefits of SCMs and DAGs -->
The structural approach to causal inference uses SCMs and DAGs to formally and graphically represent the presumed causal structure underlying the ACE [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]. Essentially, SCMs and DAGs serve as *conceptual (theoretical) models* that facilitate valid causal inference by guiding researchers in determining which statistical models can identify an estimand (ACE, CACE, or other), assuming the depicted causal structure is correct [@McElreath_2020].

SCMs and DAGs support causal inference through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Second, they depict causal relationships in a non-parametric and fully interactive way. This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp. 35]. @fig-IEflow shows the role of theoretical models in the inference process.

@@

<!-- 4. Theoretical implications of the basic building block of the SCMS and DAGs -->
A careful examination of the figures highlights the theoretical assumptions underlying the observed variables in these building blocks. SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X_{1}$ and $X_{3}$ are independent or not causally related. SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2] illustrate two connected nodes, representing a scenario where a *parent* node $X_{1}$ exerts a causal influence on a *child* node $X_{3}$. In this setup, $X_{3}$ is considered a *descendant* of $X_{1}$. Additionally, $X_{1}$ and $X_{3}$ are described as *adjacent* because there is a *direct path* connecting. SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] depict a *chain*, where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$. In this configuration, $X_{1}$ is a parent node of $X_{2}$, which is a parent node of $X_{3}$. This structure creates a *directed path* between $X_{1}$ and $X_{3}$. Consequently, $X_{1}$ is an *ancestor* of $X_{3}$, and $X_{2}$ fully *mediates* the relationship between the two. SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4] illustrate a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$. Here, $X_{2}$ is a parent node that *confounds* the relationship between $X_{1}$ and $X_{3}$. Finally, SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] show a *collider*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$. In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$ (an *inmorality*). Notably, all building blocks assume the errors are independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X1} \dsep e_{X2}$, $e_{X1} \dsep e_{X3}$, and $e_{X2} \dsep e_{X3}$.

@@

## The probabilistic implications of SCMs and DAGs {#sec-appendix-D .appendix appendix-style=plain}

<!-- 1. How the SCMS and DAGs encode probabilistic information? -->
Beyond their graphical capabilities, SCMs and DAGs can encode the probabilistic information embedded within a causal structure. They achieve this encoding by relying on three fundamental assumptions: the local Markov, the minimality, the causal edges assumptions. The *local Markov assumption* encodes probabilistic independencies between variables by declaring that nodes in a graph are independent of all its non-descendants, given its parents [@Neal_2020, pp. 20]. Meanwhile, the *minimality assumption* encodes probabilistic dependencies among variables by stating that every pair of adjacent nodes exhibits a dependency [@Neal_2020, pp. 21]. Finally, the *causal edges assumption* encodes causal relationships between variables by declaring that each parent node acts as a direct cause of its children [@Neal_2020, pp. 22]. @fig-ACflow illustrates how these assumptions influence the statistical and causal interpretations of graphs.

@@

<!-- 2. What are the general implications of this? -->
A notable implication of the assumptions underlying the probabilistic encoding is that any conceptual model described by an SCM and DAG can represent the joint distribution of variables more efficiently [@Pearl_et_al_2016, pp. 29]. This expression takes the form of a product of conditional probability distributions (CPDs) $P(child \cond parents)$. This property is formally known as the *Bayesian Network factorization* (BNF, @eq-net_factor) [@Pearl_et_al_2016, pp. 29; @Neal_2020, pp. 21]. In this expression, $pa(X_{i})$ denotes the set of variables that are the parents of $X_{i}$.

$$
P(X_{1}, X_{2}, \dots, X_{P}) = \prod^{P}_{p=1} P(X_{i} \cond pa(X_{i}) )
$$ {#eq-net_factor}

<!-- 3. What is the net result of this? -->
This encoding enables researchers with conceptual (theoretical) knowledge in the form of an SCM and DAG to predict patterns of (in)dependencies in the data. As highlighted by @Pearl_et_al_2016 [pp. 35], these predictions depend solely on the structure of these conceptual models without requiring the quantitative details of the equations or the distributions of the errors. Moreover, once researchers observe empirical data, the patterns of (in)dependencies in this data can provide significant insights into the validity of the proposed conceptual model.

@@

<!-- 4. can you illustrate this concept? -->
The five fundamental building blocks described in @sec-appendix-C clearly illustrate which (in)dependencies can SMCs and DAGs predict. For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] supports the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$). This conclusion follows from expressing their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$, which defines probabilistic independence [@Neal_2020, pp. 24]. Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \cond X_{1}) P(X_{1})$. Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in the figures of @sec-appendix-C and the CPDs.

Beyond the insights gained from two-node structures, researchers can uncover more nuanced patterns of (in)dependencies from chains, forks, and colliders. For instance, researchers applying the BNF to the chain structure depicted in the SCM [-@fig-scm_bb3] and DAG [-@fig-mdag_bb3] can represent the joint distribution for the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \cond X_{1}) P(X_{3} \cond X_{2})$. From this expression, researchers can infer additional (in)dependencies assumed in any data generated by this structure. Specifically, the structure implies that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \cond X_{2})$. In the fork structure shown in the SCM [-@fig-scm_bb4] and DAG [-@fig-mdag_bb4], researchers can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1} \cond X_{2}) P(X_{2}) P(X_{3} \cond X_{2})$. This expression allows researchers to further infer that $X_{1}$ and $X_{3}$ are unconditionally dependent $(X_{1} \ndsep X_{3})$, but conditionally independent when controlling for $X_{2}$ $(X_{1} \dsep X_{3} \cond X_{2})$. Finally, researchers analyzing the collider structure illustrated in the SCM [-@fig-scm_bb5] and DAG [-@fig-mdag_bb5] can express the joint distribution of the observed variables as $P(X_{1},X_{2},X_{3}) =$ $P(X_{1}) P(X_{2} \cond X_{1}, X_{3}) P(X_{3})$. This representation allows researchers to infer that $X_{1}$ and $X_{3}$ are unconditionally independent $(X_{1} \dsep X_{3})$, but conditionally dependent when controlling for $X_{2}$ $(X_{1} \ndsep X_{3} \cond X_{2})$. The authors @Pearl_et_al_2016 [pp. 37, 40, 41] and @Neal_2020 [pp. 25–26] provide the mathematical proofs for these conclusions.

<!-- 5. Continuing with the motivating example -->
<!-- Researchers can apply these new insights to revisit the scenario in @sec-appendix-B. From the SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1a], they can express the joint probability for the observed variables as $P(Y, T, X) = P(Y \cond T, X) P(T \cond X) P(X)$. From this expression, researchers can say that. -->

## From probability to causality {#sec-appendix-E .appendix appendix-style=plain}

<!-- 1. But, how do we identify causal effects?  -->
<!-- Researchers can use SCMs and DAGs to represent interventions of the type $P(X_{a} \cond do(X_{b}))$. These representations rely on the *modularity assumption*, which asserts that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp. 34].  -->

<!-- By leveraging these patterns, researchers can identify the ACE from associational estimates through statistical conditioning, thereby disentangling causation from mere association.  -->

<!-- Intervention relies on the concepts of Truncated factorization and manipulated graphs. The *Truncated factorization* (TF) simulates physical interventions by removing specific functions from the conceptual model and substituting them with constants, while keeping the remaining structure unchanged [Pearl_2010]. The probabilistic implications for this factorization are expressed in @eq-trunc_factor. Here, $S$ denotes the subset of variables $X_{p}$ directly impacted by the intervention. Complementary to the TF, *manipulated graphs* are DAGs with certain edges removed to represent the effects of interventions [@Neal_2020, pp. 34]. -->

$$
P(X_{1}, X_{2}, \dots, X_{P} \cond do(S)) =
\begin{cases}
  \prod P(X_{i} \cond pa(X_{i}) ) & \text{if} \: p \not\in S \\
  1 \quad & \text{otherwise}
\end{cases}
$$ {#eq-trunc_factor}

<!-- The Truncated Factorization enables researchers to define the *backdoor adjustment*. This adjustment specifies that if a variable $X_{p} \in S$ serves as a *sufficient adjustment set* for the effect of $X_{a}$ on $X_{b}$, then the ACE can be identified using @eq-backdoor [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. -->

$$
P(X_{a} \cond do(X_{b}=x)) = \sum_{Xp} P(X_{a} \cond X_{b}=x, X_{p}) P(X_{p})
$$ {#eq-backdoor}

<!-- Ultimately, backdoor adjustment enables researchers to express the ACE as: -->

$$
\begin{aligned}
\tau &= E[\tau_{i}] \\
  & = E[X_{a} \cond do(X_{b}=1)]- E[X_{a} \cond do(X_{b}=2)] \\
  & = \sum_{Xp} X_{a} \cdot P(X_{a} \cond X_{b}=1, X_{p}) P(X_{p}) - \sum_{Xp} X_{a} \cdot P(X_{a} \cond X_{b}=2, X_{p}) P(X_{p})
\end{aligned}
$$ {#eq-backdoor_adjustment}
